---
title: "Getting started with cyRAVEN"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cyRAVEN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

```{r setup}
library(cyRAVEN)
```

## What the package is for

You have a folder of files from a cytometer, one per person, and a gating
strategy written down somewhere. cyRAVEN applies that strategy to every file,
counts the cell types, compares your groups, and then spends the rest of its
effort telling you when not to believe the answer.

One principle runs through all of it, and it is worth stating first.

> The sample size is the number of people, not the number of cells.

A cytometer records until someone stops it, so one file might contain 200,000
cells and the next 4 million. If the statistics treat cells as the unit, your
sample size is really a measure of how long the machine ran, and a single donor
whose sample was recorded for longer can produce an apparently overwhelming
result on their own.

Every test here reduces each sample to one number first. Anything that cannot be
reduced that way is reported with an effect size and no p-value, which marks it
as something to follow up rather than something established.

## One call does the whole thing

```r
run_cyraven(list(
  dir              = "data/",
  recursive        = TRUE,
  outdir           = "results/",
  sample_map       = "data/sample_map.csv",
  patient_table    = "data/patient_table.csv",
  config           = system.file("config", "config_cohorts.yaml", package = "cyRAVEN"),
  group_column     = "cohort",
  reference_group  = "Healthy controls",
  max_events_per_file = 300000,
  seed             = 42
))
```

The same thing from a shell, which is how it usually gets run:

```sh
Rscript "$(Rscript -e 'cat(system.file("scripts","cyraven.R",package="cyRAVEN"))')" \
  --dir data/ --recursive --outdir results/ \
  --sample-map data/sample_map.csv \
  --group-column cohort --reference-group "Healthy controls"
```

Progress goes to stderr through `message()`, so you can silence or capture it:

```r
options(cyRAVEN.verbose = "none")     # or "inform" (default), or "debug"
```

## The stages, and why each one is there

Every stage is also a function you can call by itself. That is the reason for the
package layout: if you disagree with a step, you can replace it without forking
the pipeline.

### 1. Read

`read_fcs_resolved()` works out a marker name for each channel from `$PnS`,
falling back to `$PnN`, and applies the spillover matrix if the file carries one.
Spectral instruments write already-unmixed data with no matrix, so
`maybe_compensate()` checks and reports instead of assuming either way.
Compensating twice is as wrong as not compensating at all.

### 2. Transform

```{r cofactor}
set.seed(42)
ex <- matrix(abs(rnorm(20000 * 4, 0, 250)), ncol = 4,
             dimnames = list(NULL, c("CD3", "CD4", "CD8", "CD19")))
cf <- derive_cofactor(ex, setNames(1:4, colnames(ex)))
round(as.numeric(cf), 1)
```

The arcsinh cofactor decides how much of the near-zero region gets stretched out.
The usual default of 5 comes from mass cytometry. On spectrally unmixed flow data
it expands the background band until noise dominates every distance you care
about, so cyRAVEN derives one instead of assuming it. Each marker is bisected
until the interquartile range of its background bulk hits a target, and the panel
takes the median of those.

`derive_cofactor_pooled()` does the same across several samples rather than one,
so a single weakly stained file cannot set the scale that every threshold in the
run is then computed on.

### 3. Gate

Thresholds come from each sample's own density, not from coordinates copied
between samples:

```{r valley}
x <- c(rnorm(8000, 0, 0.4), rnorm(2000, 4, 0.5))   # a clean bimodal marker
round(density_valley(x), 2)
```

`density_valley()` returns `NA` when a marker is unimodal, which is the honest
answer. `resolve_threshold()` then tries an unstained control if one exists, and
falls back to a flagged quantile if not. Where every threshold came from is
recorded in `thresholds_used.csv`.

Per-sample thresholds are what a human gater does and they handle staining
variation correctly. They also mean a population's definition can move between
groups, which is why `stats_threshold_drift()` exists.

### 4. Score populations

Populations are a specification you write, not code:

```{r spec}
str(default_population_spec()[["CD3 pos CD4 pos"]])
```

Each entry maps a marker to `above`, `below` or `intermediate`. Editing the YAML
changes the analysis without touching any R.

### 5. Embed

```{r umap, eval = FALSE}
emb <- run_umap(mat, n_neighbors = 30, min_dist = 0.3, seed = 42)
```

`select_umap_features()` picks the features and drops scatter, viability, time
and the height and width duplicates of area channels. Keeping those was the
single biggest cause of a useless embedding in the template this package grew
out of. `plan_subsample()` then takes an equal number of cells from each sample,
so the deepest file cannot decide the shape of the manifold.

Pass `ret_model = TRUE` and later batches can be projected into the same space
with `project_umap()` instead of every coordinate moving. Asking uwot to retain
the model does change the embedding slightly, which `?run_umap` explains.

### 6. Test

```{r stats}
freq <- data.frame(
  sample_id = rep(sprintf("S%02d", 1:12), each = 2),
  population = rep(c("CD4 T cells", "B cells"), 12),
  pct_of_cd45_pos = c(rbind(rnorm(12, 25, 4), rnorm(12, 10, 2))),
  is_control = FALSE, qc_status = "pass")
grp <- setNames(rep(c("HC", "Patient"), each = 12)[1:12], sprintf("S%02d", 1:12))

res <- stats_group_comparison(freq, grp, reference = "HC")
res[, c("population", "comparison_group", "n_reference", "cliffs_delta", "p_value")]
```

`stats_marker_state()` does the same for marker intensity inside a population,
and tests both the median and the fraction positive, because a shift that affects
only part of a population moves one of those and not the other.

### 7. Diagnose

This is the stage that decides how much of the rest you should believe. The
[diagnostics article](diagnostics.html) covers all of it; in short:

`batch_mixing_report()` measures how strongly acquisition batch structures the
embedding, against a permutation null, and separately how far batch overlaps with
your study group. Correction is available through `--correct-batch` and is
refused when that overlap is too large, because past a point removing the batch
and removing your finding are the same operation.

`stats_threshold_drift()` asks whether a marker's gate moves systematically
between groups, which would make part of any abundance difference definitional
rather than biological.

`stats_confounding()` asks whether age or sex is both unbalanced across groups
and associated with the outcome. Those two facts together are what make something
a confounder.

`cluster_gate_agreement()` compares unsupervised clusters against your gate
labels. It is the only output that can find a population your specification never
described, or show you that a gate is rejecting cells that obviously belong to
it.

### 8. Explain

`cluster_gate_agreement()` can tell you a cluster matches nothing you named. It
cannot tell you what the cluster is, which leaves you holding a cluster number
and a cell count and nothing you can take back to the instrument.

`explain_cluster()` runs the gating stage backwards to close that gap. Given a
set of labelled cells it picks the two most discriminating markers, fits a convex
polygon in that plane, keeps the cells inside it, and repeats. That is the shape
of a manual gating strategy, which is why the answer can be drawn by hand.

```r
st <- explain_cluster(marker_matrix, cluster_id == 7)
st$summary        # marker pair and metrics per level
st$levels[[1]]$polygon   # vertices, in arcsinh units
```

A polygon rather than more thresholds, because a conjunction of one-dimensional
cuts is an axis-aligned rectangle, and a rectangle laid over a diagonal boundary
has to either let contaminants in or throw real cells away.
`derive_singlet_band()` is already a hand-written instance of that same problem.

Two things to keep in mind when reading the output. Every metric is computed on
cells held out of the fit, with the in-sample value printed next to it, because
eight free half-planes can memorise a few thousand cells and the gap between
those two numbers tells you whether that happened. And the whole stage is
descriptive. It says where cells sit in marker space, not that they are a real
population. It carries no p-value, and nothing it produces feeds back into the
analysis.

### 9. Report

Figures, CSVs, and `write_run_manifest()`, which records the R version, package
versions, the git commit and whether the checkout was dirty, the full
invocation, and every input file with its size and modification time. It is
written before the expensive stages and rewritten on the way out, so a run that
crashes leaves a manifest saying `status: failed` rather than leaving nothing.

## Where to go next

- [Gating and populations](gating.html), for how to write your own specification.
- [Checking the result](diagnostics.html), for everything that can tell you the
  gates were wrong.
- [Statistics](statistics.html), for why these tests and not others.
- [Using cyRAVEN with cyCONDOR](with-cycondor.html), if you want a
  clustering-first view alongside this one.
- `?run_cyraven` for the full option list, and
  `system.file("examples", package = "cyRAVEN")` for template metadata files.
