---
title: "Getting started with cyRAVEN"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with cyRAVEN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

```{r setup}
library(cyRAVEN)
```

## What this package is for

`cyRAVEN` takes a directory of FCS files from one experiment and produces a
gated, embedded, statistically tested result set: population frequencies per
sample, a shared UMAP per marker panel, differential abundance and differential
state between cohorts, and a set of diagnostics whose job is to tell you when not
to believe the rest.

It is built around one opinion, which is worth stating before anything else:

> **The replicates are the donors, not the cells.**

A flow cytometry file contains as many events as somebody chose to acquire. Using
cells as the unit of analysis makes the p-value a function of acquisition time,
so a marker difference driven entirely by one deeply-acquired donor arrives
looking overwhelming. Every test in this package therefore aggregates to one
value per sample first, and every quantity that *cannot* be aggregated that way
is labelled descriptive and reported without a p-value.

## The whole pipeline in one call

```r
run_cyraven(list(
  dir              = "data/",
  recursive        = TRUE,
  outdir           = "results/",
  sample_map       = "data/sample_map.csv",
  patient_table    = "data/patient_table.csv",
  config           = system.file("config", "config_cohorts.yaml", package = "cyRAVEN"),
  group_column     = "cohort",
  reference_group  = "Healthy controls",
  max_events_per_file = 300000,
  seed             = 42
))
```

The same thing from a shell, which is how it is usually run:

```sh
Rscript "$(Rscript -e 'cat(system.file("scripts","cyraven.R",package="cyRAVEN"))')" \
  --dir data/ --recursive --outdir results/ \
  --sample-map data/sample_map.csv \
  --group-column cohort --reference-group "Healthy controls"
```

Progress goes to `stderr` through `message()`, so it can be silenced or captured:

```r
options(cyRAVEN.verbose = "none")     # or "inform" (default), or "debug"
```

## The nine stages, and why each exists

Every stage is also a function you can call on its own. That is the point of the
package layout: a step you disagree with can be replaced without forking the
pipeline.

### 1. Read

`read_fcs_resolved()` resolves each channel to a marker symbol from `$PnS`,
falling back to `$PnN`, and applies the spillover matrix when the file carries
one. Spectral instruments write already-unmixed data with no matrix, so
`maybe_compensate()` detects and reports rather than assuming either way —
compensating twice is as wrong as not compensating.

### 2. Transform

```{r cofactor}
set.seed(42)
ex <- matrix(abs(rnorm(20000 * 4, 0, 250)), ncol = 4,
             dimnames = list(NULL, c("CD3", "CD4", "CD8", "CD19")))
cf <- derive_cofactor(ex, setNames(1:4, colnames(ex)))
round(as.numeric(cf), 1)
```

The arcsinh cofactor decides how much of the near-zero region gets expanded. The
conventional default of 5 comes from mass cytometry; on spectrally unmixed flow
data it expands the background band until noise dominates every distance. So it
is derived rather than assumed: each marker is bisected until the interquartile
range of its background bulk hits a target, and the panel takes the median.

`derive_cofactor_pooled()` does this across several samples rather than one,
because a single weakly-stained file should not set the scale on which every
threshold in the run is computed.

### 3. Gate

Thresholds come from each sample's own density, not from coordinates copied
between samples:

```{r valley}
x <- c(rnorm(8000, 0, 0.4), rnorm(2000, 4, 0.5))   # a clean bimodal marker
round(density_valley(x), 2)
```

`density_valley()` returns `NA` when a marker is unimodal — the honest answer,
which then routes `resolve_threshold()` to an unstained control if one exists,
and to a flagged quantile fallback if not. The source of every threshold is
recorded in `thresholds_used.csv`.

Per-sample thresholds are what a human gater does and adapt correctly to staining
variation. They also mean a population's *definition* can move between cohorts,
which is why `stats_threshold_drift()` exists — see the diagnostics section.

### 4. Score populations

Populations are a declarative specification, not code:

```{r spec}
str(default_population_spec()[["CD3 pos CD4 pos"]])
```

Each entry maps a marker to `above`, `below` or `intermediate`, with `any_of` for
genuine disjunctions. Editing the YAML changes the analysis without touching R.

### 5. Embed

```{r umap, eval = FALSE}
emb <- run_umap(mat, n_neighbors = 30, min_dist = 0.3, seed = 42)
```

Features are chosen by `select_umap_features()`, which drops scatter, viability,
time and the height/width duplicates of area channels — retaining those was the
single largest cause of an uninformative embedding in the original template.
Cells are subsampled equally per sample by `plan_subsample()` so the deepest file
cannot dictate the manifold.

Pass `ret_model = TRUE` and later batches can be projected into the same space
with `project_umap()` instead of moving every coordinate. Note that asking uwot to
retain the model changes the embedding — see `?run_umap`.

### 6. Test

```{r stats}
freq <- data.frame(
  sample_id = rep(sprintf("S%02d", 1:12), each = 2),
  population = rep(c("CD4 T cells", "B cells"), 12),
  pct_of_cd45_pos = c(rbind(rnorm(12, 25, 4), rnorm(12, 10, 2))),
  is_control = FALSE, qc_status = "pass")
grp <- setNames(rep(c("HC", "Patient"), each = 12)[1:12], sprintf("S%02d", 1:12))

res <- stats_group_comparison(freq, grp, reference = "HC")
res[, c("population", "comparison_group", "n_reference", "cliffs_delta", "p_value")]
```

`stats_marker_state()` does the same for marker intensity inside a population,
testing both the median and the fraction positive because a shift affecting a
subset moves one and not the other.

### 7. Diagnose

The stage that decides how much of the rest to believe.

- `batch_mixing_report()` measures how much acquisition batch structures the
  embedding, against a permutation null, **and** how far batch is confounded with
  cohort. The package deliberately does not correct batch effects: where batch and
  cohort overlap, removing the batch and removing the finding are the same
  operation.
- `stats_threshold_drift()` asks whether a marker's gate moves systematically
  between cohorts, which would make part of any abundance difference definitional.
- `stats_confounding()` asks whether age or sex is both unbalanced across cohorts
  and associated with the outcome — the two facts that together make a confounder.
- `cluster_gate_agreement()` compares unsupervised clusters against the gate
  labels, and is the only output that can find a population the specification does
  not describe, or show that a gate is rejecting cells that plainly belong to it.

### 8. Explain

`cluster_gate_agreement()` can report that a cluster matches no described
population. It cannot say what that cluster *is*, which leaves the finding as a
cluster number and a cell count — nothing you can take back to the instrument.

`explain_cluster()` runs the gating stage backwards to close that loop. Given a
set of labelled cells it picks the two most discriminating markers, fits a
**convex polygon** gate in that plane, keeps the cells inside, and repeats — the
shape of a manual gating strategy, which is why the result can be executed by
hand.

```r
st <- explain_cluster(marker_matrix, cluster_id == 7)
st$summary        # marker pair and metrics per level
st$levels[[1]]$polygon   # vertices, in arcsinh units
```

A polygon rather than more thresholds because a conjunction of one-dimensional
cuts is an axis-aligned rectangle, and a rectangle over a diagonal boundary must
either admit contaminants or reject real cells. `derive_singlet_band()` is
already a hand-written instance of exactly that problem.

Two things to hold onto when reading the output. Every metric is computed on
cells **held out of the fit**, with the in-sample value beside it — eight free
half-planes can memorise a few thousand cells, so the gap between the two numbers
is information. And the whole stage is descriptive: it says where cells sit in
marker space, not that they are a real population, it carries no p-value, and
nothing it produces re-enters the analysis.

### 9. Report

Figures and CSVs, plus `write_run_manifest()` — R version, package versions, git
commit and whether the checkout was dirty, the full invocation, and every input
file with its size and mtime. It is written *before* the expensive stages and
rewritten on the way out, so a crashed run leaves `status: failed` rather than
nothing.

## Where to go next

- `vignette("statistics", package = "cyRAVEN")` — why these tests and not others.
- `?run_cyraven` for the full option list.
- `system.file("examples", package = "cyRAVEN")` for template metadata files.
