---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
library(cyRAVEN)
```

## 1. Scope

cyRAVEN quantifies populations declared in advance across a batch of FCS files
and evaluates whether the declaration held. The pipeline is a single call; every
stage is also an exported function callable in isolation, so a step can be
replaced without forking the package.

One constraint governs the design and is stated before the stages.

> The unit of replication is the donor, not the event.

Event counts are set by acquisition duration rather than by study design.
Inference over events sets the degrees of freedom from instrument time and allows
one deeply acquired donor to determine a group result. Every test aggregates to
one value per sample before estimation. Quantities that cannot be aggregated are
reported with effect sizes and no p-value.

## 2. Invocation

The supported execution path is the container. `uwot` is a stochastic embedder
whose output varies with its own version and with the BLAS beneath it, and every
gate is placed at a kernel density minimum, so a different density implementation
moves thresholds and the frequencies derived from them. The image pins R 4.4.3, a
dated CRAN snapshot and Bioconductor 3.20.

A run takes the FCS files, one sample sheet and one config. Both files are
described in the [Inputs article](inputs.html).

### 2.1 First run on a new cohort

The sequence, in order. Each step is cheap and catches what the next one would
otherwise waste time on.

```sh
# 0. build the image once
docker build -f inst/scripts/Dockerfile -t cyraven:1.0.0 .

mkdir -p data results

# 1. a sheet covering every file in the directory
docker run --rm -v "$PWD/data:/data" cyraven:1.0.0 \
  --dir /data/fcs --recursive --write-samples /data/samples.csv

# 2. edit data/samples.csv: patient_id, cohort, and any study variable

# 3. start the config from the shipped template
docker run --rm --entrypoint sh cyraven:1.0.0 -c \
  'cat /usr/local/lib/R/site-library/cyRAVEN/examples/analysis_template.yaml' \
  > data/analysis.yaml

# 4. validate both files against the FCS headers, in seconds
docker run --rm -v "$PWD/data:/data:ro" -v "$PWD/results:/results" \
  cyraven:1.0.0 --dir /data/fcs --recursive \
  --samples /data/samples.csv --config /data/analysis.yaml \
  --outdir /results --check

# 5. edit populations: in the config until --check reports no problems,
#    then run
docker run --rm -v "$PWD/data:/data:ro" -v "$PWD/results:/results" \
  cyraven:1.0.0 --dir /data/fcs --recursive \
  --samples /data/samples.csv --config /data/analysis.yaml \
  --group-column cohort --reference-group "Healthy controls" \
  --outdir /results

# 6. read results/report.html
```

On Windows PowerShell substitute `${PWD}` for `$PWD`; on Git Bash prefix with
`MSYS_NO_PATHCONV=1`, or the shell rewrites `/data` into a Windows path.

### 2.2 The full analysis

Batch diagnostics and the unsupervised check are opt-in because each adds
outputs and runtime.

```sh
docker run --rm -v "$PWD/data:/data:ro" -v "$PWD/results:/results" \
  cyraven:1.0.0 \
  --dir /data/fcs --recursive \
  --samples /data/samples.csv --config /data/analysis.yaml \
  --group-column cohort --reference-group "Healthy controls" \
  --batch-column acquisition_date --cluster \
  --covariates age,sex --reference-date 2025-01-01 \
  --no-session \
  --outdir /results
```

`--no-session` is worth adding on a large cohort: `session_state.RData` can reach
several hundred MB and dominate the runtime, and no result depends on it.

### 2.3 A large cohort that must fit in memory

Peak memory is set by the events held at once, not by the number of files.

```sh
docker run --rm -v "$PWD/data:/data:ro" -v "$PWD/results:/results" \
  cyraven:1.0.0 \
  --dir /data --recursive \
  --samples /data/samples.csv --config /data/analysis.yaml \
  --max-events-per-file 300000 \
  --cells-per-sample 6000 --max-cells 150000 \
  --no-session --outdir /results
```

Subsampling raises every detection limit in proportion, because it lowers the
parent-gate count each frequency is a fraction of.

Run one analysis at a time. Peak memory is dominated by the embedding, which
holds `--max-cells` cells across every marker, and two concurrent runs at these
settings will exhaust a workstation. The stages are already multi-threaded, so
nothing is gained by running containers in parallel. A run killed by the
out-of-memory killer exits 137 and writes no report, because SIGKILL runs no
handler; that combination is the signature to look for.

### 2.4 Comparing a run against an accepted one

```sh
# once, on the run being adopted as the reference
docker run --rm -v "$PWD/data:/data:ro" -v "$PWD/results:/results" \
  cyraven:1.0.0 --dir /data/fcs --samples /data/samples.csv \
  --config /data/analysis.yaml --outdir /results \
  --write-baseline /results/baseline.rds

# on every later run
docker run --rm -v "$PWD/data:/data:ro" -v "$PWD/results2:/results" \
  -v "$PWD/results/baseline.rds:/baseline.rds:ro" \
  cyraven:1.0.0 --dir /data/fcs --samples /data/samples.csv \
  --config /data/analysis.yaml --outdir /results \
  --baseline /baseline.rds --fail-on-drift
```

`--fail-on-drift` exits non-zero after every output has been written, so it can
gate a pipeline without costing the diagnostics that explain the failure.

### 2.5 From an R session

Where the package is installed locally and bit-level reproducibility against the
pinned environment is not required:

```r
library(cyRAVEN)

run_cyraven(list(
  dir                 = "data/fcs",
  recursive           = TRUE,
  outdir              = "results/",
  samples             = "data/samples.csv",
  config              = "data/analysis.yaml",
  group_column        = "cohort",
  reference_group     = "Healthy controls",
  max_events_per_file = 300000,
  seed                = 42
))
```

Validate first, exactly as `--check` does from the command line:

```r
run_cyraven(list(dir = "data/fcs", samples = "data/samples.csv",
                 config = "data/analysis.yaml", outdir = "results/",
                 check_only = TRUE))
```

Both forms call the same `run_cyraven()`. The command-line front end resolves to
`system.file("scripts", "cyraven.R", package = "cyRAVEN")` and is what the
container entrypoint executes.

Progress is emitted through `message()` and directed to stderr:

```r
options(cyRAVEN.verbose = "none")     # or "inform" (default), or "debug"
```

A run that stops on an error still writes `report.html`, carrying the error, the
stage it reached, the log leading up to it, what the message means for the data,
the next action, and every output produced before the failure.

## 3. Stages

### 3.1 Read

`read_fcs_resolved()` resolves channels to marker symbols from `$PnS` with
fallback to `$PnN`, and applies the acquisition spillover matrix where the file
carries one. Spectral instruments write unmixed data with no matrix, so
`maybe_compensate()` detects and reports rather than assuming; applying
compensation twice is as damaging as omitting it.

### 3.2 Transform

```{r cofactor}
set.seed(42)
ex <- matrix(abs(rnorm(20000 * 4, 0, 250)), ncol = 4,
             dimnames = list(NULL, c("CD3", "CD4", "CD8", "CD19")))
cf <- derive_cofactor(ex, setNames(1:4, colnames(ex)))
round(as.numeric(cf), 1)
```

The arcsinh cofactor governs the width of the quasi-linear region near zero. The
conventional value of 5 derives from mass cytometry and over-expands the
background band on spectrally unmixed fluorescence data. Each marker is bisected
until the interquartile range of its background distribution reaches a target,
and the panel adopts the median. `derive_cofactor_pooled()` estimates across
samples so that a single weakly stained file does not set the scale on which
every threshold in the run is computed.

### 3.3 Gate

```{r valley}
x <- c(rnorm(8000, 0, 0.4), rnorm(2000, 4, 0.5))   # bimodal marker
round(density_valley(x), 2)
```

Thresholds are placed at the density minimum within each sample.
`density_valley()` returns `NA` for unimodal distributions, after which
`resolve_threshold()` attempts a declared unstained control and falls back to a
flagged quantile. Derivation is recorded per threshold in `thresholds_used.csv`.

Per-sample thresholding accommodates staining variation but permits the
definition of a population to differ between groups, which
`stats_threshold_drift()` exists to detect.

### 3.4 Score

```{r spec}
str(default_population_spec()[["CD3 pos CD4 pos"]])
```

Populations are evaluated as Boolean conjunctions of marker directions read from
YAML. Editing the specification changes the analysis without modifying R code.

### 3.5 Quantify the uncertainty

```{r unc, eval = FALSE}
u <- threshold_uncertainty(x_parent, source = "valley", B = 100)
c(u$u_sampling, u$u_method, u$u_combined)
```

Each cut is re-derived from resamples of its own events, and again across the
settings that placed it. The two spreads combine in quadrature and propagate to
every population reading that marker, alongside the CD45 parent term, which
enters all of them because it fixes the denominator.

The reported threshold is unchanged: the perturbation runs on copies. Every entry
point saves and restores `.Random.seed`, so the cell selection in the next stage
draws from the same stream position whether this runs or not.

### 3.6 Embed

```{r umap, eval = FALSE}
emb <- run_umap(mat, n_neighbors = 30, min_dist = 0.3, seed = 42)
```

`select_umap_features()` excludes scatter, viability, time and the height and
width duplicates of area channels; retaining these was the principal cause of
uninformative embeddings in the precursor implementation. `plan_subsample()`
draws equal cell numbers per sample so that acquisition depth does not determine
the manifold.

`ret_model = TRUE` retains the model for projection of later batches through
`project_umap()`, which holds coordinates fixed between runs. Model retention
alters the embedding; see `?run_umap`.

### 3.7 Test

```{r stats}
freq <- data.frame(
  sample_id = rep(sprintf("S%02d", 1:12), each = 2),
  population = rep(c("CD4 T cells", "B cells"), 12),
  pct_of_cd45_pos = c(rbind(rnorm(12, 25, 4), rnorm(12, 10, 2))),
  is_control = FALSE, qc_status = "pass")
grp <- setNames(rep(c("HC", "Patient"), each = 12)[1:12], sprintf("S%02d", 1:12))

res <- stats_group_comparison(freq, grp, reference = "HC")
res[, c("population", "comparison_group", "n_reference", "cliffs_delta", "p_value")]
```

`stats_marker_state()` applies the same treatment to marker expression within a
population, testing median intensity and percent positive separately because a
shift confined to a subset moves one and not the other.

### 3.8 Diagnose

Six outputs constrain interpretation of the preceding stages: threshold drift,
gate and counting uncertainty, phenotype concordance, gate against cluster
concordance, learned gate geometry, and conformance against a baseline. The
[diagnostics article](diagnostics.html) covers each in full.

`batch_mixing_report()` quantifies batch structure in the embedding against a
permutation null and reports Cramér's *V* between batch and group. Correction
through `--correct-batch` is refused above a configurable *V*, where batch and
biological effect are not separable.

`stats_threshold_drift()` tests whether marker thresholds separate by group,
which would render an abundance difference partly definitional.

`stats_confounding()` reports whether a covariate both differs across groups and
associates with the outcome, the two conditions jointly defining a confounder.

`cluster_gate_agreement()` cross-tabulates declared labels against unsupervised
clusters and is the only output capable of identifying a population absent from
the specification, or a threshold rejecting events that belong to its population.

`run_gate_uncertainty()` reports how far each threshold moves under resampling and
propagates that to every population reading it, so a difference smaller than its
own gate's movement is identifiable as such.

### 3.9 Explain

`cluster_gate_agreement()` identifies a cluster matching no declared population
but cannot characterise it, leaving a cluster index and a cell count.

`explain_cluster()` inverts the gating stage. Given labelled events it selects
the two most discriminating markers, fits a convex polygon in that plane, retains
the enclosed events and recurses, producing the topology of a manual gating
strategy.

```r
st <- explain_cluster(marker_matrix, cluster_id == 7)
st$summary               # marker pair and metrics per level
st$levels[[1]]$polygon   # vertices, in arcsinh units
```

Convex polygons rather than thresholds: a conjunction of one-dimensional cuts is
an axis-aligned rectangle, and an oblique boundary constrained to a rectangle
must admit contaminants or exclude genuine events. `derive_singlet_band()` is a
hand-written instance of the same geometry.

Metrics are computed on held-out events with the resubstitution value alongside,
since eight free half-planes can memorise a few thousand events. The stage is
descriptive: it carries no p-value and scored frequencies are identical whether
it runs or not.

`--external-labels` supplies the labels from elsewhere, a cyCONDOR clustering for
instance, and `gate_transferability()` then refits the strategy with one donor
withheld and scores it on that donor. Held-out events come from the same donors
and overstate how well a gate travels; the per-donor minimum does not.
`--export-gates` writes the result as Gating-ML 2.0 in the units the FCS file
stores.

### 3.10 Report

Figures, tables, and `write_run_manifest()`, which records R version, package
versions, git commit and working-tree state, the full invocation, and every input
file with size and modification time. The manifest is written before the
expensive stages and rewritten at completion, so an interrupted run records
`status: failed`.

## 4. Further reading

- [Gating](gating.html), for threshold derivation and specification syntax.
- [Diagnostics](diagnostics.html), for the outputs that falsify a specification.
- [Statistics](statistics.html), for test selection and its assumptions.
- [Interoperability](with-cycondor.html), for combined use with cyCONDOR.
- `?run_cyraven` for the option list;
  `system.file("examples", package = "cyRAVEN")` for metadata templates.
