---
title: "Statistical choices, and what they cost"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical choices, and what they cost}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

```{r setup}
library(cyRAVEN)
```

This vignette explains why the tests are the ones they are. Every choice below
trades something away, and the trade is stated rather than implied.

## 1. Cells are not replicates

The number of cells in an FCS file is a property of how long somebody left the
cytometer running. Treating cells as independent observations makes the p-value a
function of acquisition time.

Concretely: pooling cells across a cohort and running a rank test on marker
intensity gives *n* in the tens of thousands. Almost any difference reaches
significance, and a difference produced entirely by one deeply-acquired donor is
indistinguishable from one shared by the whole cohort.

The remedy is the one `diffcyt` uses: aggregate to one value per sample per
population first, then model those values. `stats_marker_state()` does exactly
this, and *n* becomes the number of donors — which is what the design actually
provides.

```{r ds}
mfi <- expand.grid(sample_id = sprintf("S%02d", 1:12),
                   population = "CD4 T cells", marker = "HLA-DR",
                   stringsAsFactors = FALSE)
mfi$n_cells <- 500
mfi$median_asinh <- rnorm(12, 2, 0.3) + rep(c(0, 0.8), each = 6)
mfi$pct_positive <- pmin(100, pmax(0, rnorm(12, 30, 5) + rep(c(0, 10), each = 6)))
mfi$is_control <- FALSE; mfi$qc_status <- "pass"
grp <- setNames(rep(c("HC", "Patient"), each = 6), sprintf("S%02d", 1:12))

res <- stats_marker_state(mfi, grp, reference = "HC")
res[, c("marker", "measure", "n_reference", "n_comparison", "delta_median", "p_value")]
```

Note `n_reference` is 6, not 3,000.

The package still reports cell-level effect sizes — `stats_subcluster_shifts()`
ranks marker shifts over pooled cells — but deliberately attaches **no p-value**
to them. They are for generating hypotheses, not testing them.

## 2. Rank tests, not limma or a mixed model

`diffcyt` fits limma's moderated *t*, borrowing variance across markers through
an empirical-Bayes prior. That is an excellent trade at mass-cytometry marker
counts (30–40) and designed-experiment sample counts.

At the sizes this package is built for — a dozen markers, single-digit donors per
cohort — the prior is estimated from too few markers to stabilise anything, and
the normality assumption stops being a formality and starts doing real work.
Wilcoxon and Kruskal-Wallis assume only independence and ordinal comparability.

**What that costs:** less power when the data really are normal, and no clean way
to include covariates. Both are accepted deliberately.

**When to switch:** roughly 15+ samples per cohort. At that point the prior earns
its keep and a mixed model can absorb repeated measures properly.

## 3. Percentages are compositional

Population percentages within a sample sum to 100. They cannot move
independently: a granulocyte expansion mechanically depresses every lymphocyte
percentage, and those depressions test significant with no lymphocyte having
changed.

`clr_frequencies()` applies the centred log-ratio, expressing each population
relative to the geometric mean of that sample's composition.

```{r clr}
freq <- data.frame(
  sample_id = rep(sprintf("S%02d", 1:6), each = 3),
  population = rep(c("Granulocytes", "CD4 T cells", "B cells"), 6),
  pct_of_cd45_pos = as.vector(replicate(6, {v <- c(60, 25, 15) * exp(rnorm(3, 0, .1)); 100*v/sum(v)})),
  is_control = FALSE, qc_status = "pass")
head(clr_frequencies(freq)[, c("sample_id", "population", "pct_of_cd45_pos", "clr")], 3)
```

Both the raw and the CLR test are reported, because the **comparison between
them** is the finding: `compositional_concordance()` labels each result
`robust_to_composition`, `raw_only__possible_composition_artefact`, or
`clr_only__was_masked_by_composition`.

**What CLR does not fix:** it cannot recover absolute cell numbers. If every
population doubles, the composition is unchanged and CLR sees nothing. Only
cells/µL distinguishes "expanded" from "expanded relative to the rest", which is
why `abundance_measure()` prefers absolute counts whenever the patient table
supplies the white-cell count. **CLR is the right test for frequency data;
absolute counts are better data.**

## 4. Confounders are diagnosed, not adjusted away

The patient table carries age and sex. It is tempting to put them in a model.

At single-digit *n* per cohort, that model spends most of its residual degrees of
freedom on nuisance terms — and if age is strongly associated with cohort, which
in a syndrome-versus-adult-control design it usually is, the two are not separable
at *any* sample size. The adjusted estimate is extrapolation, and it arrives
wearing a confidence interval.

So `stats_confounding()` reports the two facts that *together* make something a
confounder: whether the covariate differs across cohorts, and whether it tracks
the outcome. Either alone is harmless. `stats_rank_ancova()` will fit the adjusted
model on request, labels every row `EXPLORATORY`, and writes `NOT FITTED` with a
reason rather than a number when the residual degrees of freedom are insufficient.

## 5. Batch effects are measured, not removed

Harmony and CytoNorm both move cells until batch labels mix. When batch is
confounded with the biological group — samples of one cohort acquired on one set
of days, which is how clinical cohorts are normally collected — "removing the
batch" and "removing the effect you are looking for" are the same operation, and
no algorithm can tell them apart.

`batch_mixing_report()` therefore returns two numbers: iLISI against a permutation
null (is there batch structure?) and Cramér's V (can it even be separated from
cohort?). If V is high, no correction is safe at any setting, and that is the most
useful thing the diagnostic can say.

## 6. Multiplicity

Benjamini-Hochberg within each test family. Differential state adjusts *within*
each measure rather than across both: the median and the fraction-positive are two
views of the same cells and are strongly correlated, so pooling them would inflate
the family with tests that are not independent questions and over-penalise every
hit.

## 7. Effect sizes

Every p-value is reported next to an effect size — Cliff's delta over samples, or
matched-pairs rank-biserial. Differences, never ratios, for arcsinh values: that
scale runs negative, where a ratio is uninterpretable and flips sign for no
biological reason.

## Summary

| Question | Function | Unit |
|---|---|---|
| Did a population change size? | `stats_group_comparison()` | sample |
| …after removing the compositional constraint? | `clr_frequencies()` + `compositional_concordance()` | sample |
| Did a marker change level inside a population? | `stats_marker_state()` | sample |
| Is it really age or sex? | `stats_confounding()` | sample |
| Is it really the batch? | `batch_mixing_report()` | cell (descriptive) |
| Is it really the gate? | `stats_threshold_drift()`, `cluster_gate_agreement()` | sample / cell |
| Which markers shift within a subcluster? | `stats_subcluster_shifts()` | cell (no p-value) |
