---
title: "Statistical choices, and what they cost"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical choices, and what they cost}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

```{r setup}
library(cyRAVEN)
```

Every statistical test involves a choice, and every choice gives something up.
This page says what each one gives up, so you can judge whether you agree.

## Count people, not cells

This is the most important idea here, and it costs nothing to understand.

Suppose you study eight patients and eight controls. Your sample size is 16
people. But each blood sample contains a million cells, so if you run a test
across all the cells, the software thinks your sample size is 16 million.

That is not a technicality. Two things go wrong.

First, the number of cells depends on how long the operator left the machine
running, so your sample size is really a measure of laboratory scheduling. Run one
patient's tube for an extra ten minutes and your study appears more powerful.

Second, and worse, one person can produce the whole result. If a single patient's
sample was recorded for longer than everyone else's, their cells outnumber the
rest and their personal quirks look like a finding about the disease.

cyRAVEN reduces every sample to a single number for each cell type before any
test runs, so the sample size stays at 16. That is the same approach the diffcyt
method uses, and it is the reason the numbers below look modest rather than
impressive.

```{r ds}
mfi <- expand.grid(sample_id = sprintf("S%02d", 1:12),
                   population = "CD4 T cells", marker = "HLA-DR",
                   stringsAsFactors = FALSE)
mfi$n_cells <- 500
mfi$median_asinh <- rnorm(12, 2, 0.3) + rep(c(0, 0.8), each = 6)
mfi$pct_positive <- pmin(100, pmax(0, rnorm(12, 30, 5) + rep(c(0, 10), each = 6)))
mfi$is_control <- FALSE; mfi$qc_status <- "pass"
grp <- setNames(rep(c("HC", "Patient"), each = 6), sprintf("S%02d", 1:12))

res <- stats_marker_state(mfi, grp, reference = "HC")
res[, c("marker", "measure", "n_reference", "n_comparison", "delta_median", "p_value")]
```

Look at `n_reference`: it is 6, the number of control donors, not 3,000, the
number of cells they contributed.

Cell-level numbers are still produced, in `stats_subcluster_shifts()`, but they
carry an effect size and deliberately no p-value. They are there to suggest what
to look at next, not to establish anything.

## Tests that do not assume a bell curve

The tests here are Wilcoxon and Kruskal-Wallis, which work on the rank order of
values rather than their exact size. They assume only that your samples are
independent and that the values can be put in order.

The main alternative, used by diffcyt, is a method that assumes the data follows
a normal distribution and borrows information across proteins to stabilise its
estimates. With 30 to 40 proteins and a properly sized study, that is the better
choice and gives more power.

With a dozen proteins and eight people per group, it stops being the better
choice. There are too few proteins for the borrowing to help, and the normality
assumption starts doing real work rather than being a formality, at exactly the
sample size where you cannot check whether it holds.

What the rank tests cost you: some power when the data really are normal, and no
tidy way to adjust for age or sex.

When to switch: around fifteen or more samples per group.

## Percentages are a fixed pie

Every cell type is reported as a percentage of white blood cells, and those
percentages must add to 100. That means they cannot move independently, and it
produces a trap that catches people constantly.

Suppose a patient group has a genuine expansion of granulocytes, which are the
most common white cell. Because the total is fixed at 100%, every other cell type
must fall as a percentage. Your T cells, B cells and NK cells will all appear
reduced and may all test significant, when in absolute terms not one of them has
changed. The patient has the same number of T cells per millilitre of blood as
everybody else.

cyRAVEN runs each comparison twice: once on the raw percentages, and once after a
transformation that expresses each population relative to the sample's overall
composition rather than to a fixed total.

```{r clr}
freq <- data.frame(
  sample_id = rep(sprintf("S%02d", 1:6), each = 3),
  population = rep(c("Granulocytes", "CD4 T cells", "B cells"), 6),
  pct_of_cd45_pos = as.vector(replicate(6, {v <- c(60, 25, 15) * exp(rnorm(3, 0, .1)); 100*v/sum(v)})),
  is_control = FALSE, qc_status = "pass")
head(clr_frequencies(freq)[, c("sample_id", "population", "pct_of_cd45_pos", "clr")], 3)
```

Comparing the two answers is where the value is, and
`compositional_concordance()` does that for you, labelling every result as one of
three things.

`robust_to_composition` means the difference held up both ways, and you can
believe it.

`raw_only__possible_composition_artefact` means it was significant on raw
percentages but vanished once composition was accounted for. This is the
granulocyte trap above. Something else in the sample probably moved.

`clr_only__was_masked_by_composition` is the reverse: a real change that the
fixed total was hiding.

One thing this cannot fix. If every cell type in a patient doubles, the
percentages are all unchanged and no amount of clever transformation will notice,
because the information is simply not in a percentage. Only an actual count of
cells per microlitre of blood distinguishes "this population grew" from "this
population grew relative to the others". If your patient table has a white cell
count from a routine blood test, cyRAVEN will use it and give you real counts.
Percentages are what you work with when you have nothing better; counts are
better data.

## Age and sex are flagged, not silently corrected

Your patient table may have age and sex in it, and the instinct is to put them
into the model and let the statistics handle it.

That works when you have enough people. With eight per group it does not. The
model has to estimate an age effect and a sex effect out of the same 16
observations you wanted for the main comparison, leaving very little for the
question you care about.

There is a harder version of the problem. If your patients are all children and
your controls are all adults, there is no way to separate age from disease,
because you have no young controls and no old patients to learn the difference
from. No sample size fixes this. Any adjusted number the model produces is an
extrapolation into a region where you have no data, and it will arrive with a
confidence interval that makes it look trustworthy.

So `stats_confounding()` reports the facts and lets you judge: does age differ
between groups, and is age related to the population you are measuring. Both have
to be true before there is a problem. If you still want the adjusted model,
`stats_rank_ancova()` will fit it, mark every row `EXPLORATORY`, and write
`NOT FITTED` with a reason wherever there is not enough data to support it.

## Batch effects are measured before anything is removed

The standard tools for this shift cells around until samples from different
processing days overlap. That works when the days are unrelated to the study
design.

Clinical studies rarely oblige. Patients get recruited and run in one period,
controls in another, so "which day it was processed" and "patient or control"
become close to the same variable. Shifting cells until the days overlap then
also shifts away the difference you are studying, and the method has no way to
tell the two apart.

`batch_mixing_report()` answers both questions before anything is changed: is
there a batch effect, judged against what random labelling would look like, and
can it be told apart from the study group at all. If the answer to the second is
no, then no setting makes correction safe, and saying so is the most useful thing
it can do.

Correction is available with `--correct-batch` and is refused above a threshold
you control. The refusal is the point of the feature, not an obstacle to it.

## Testing many things at once

Test twenty populations at the usual 5% threshold and you expect one false
positive even when nothing is going on. cyRAVEN applies the Benjamini-Hochberg
correction, which adjusts the p-values for how many tests were run, and reports
both the raw and adjusted values.

One detail. For marker brightness, cyRAVEN measures two things per population:
the median level, and the percentage of cells that are positive. These are two
views of the same cells and tend to move together, so the correction is applied
within each measure rather than pooling them. Pooling would double the apparent
number of independent questions and penalise every real result for it.

## Effect sizes sit next to every p-value

A p-value tells you whether a difference is likely to be real. It says nothing
about whether it is big enough to matter. Every test here reports an effect size
alongside, so you can see both.

For transformed brightness values, cyRAVEN reports differences rather than fold
changes. Those scales run below zero, and a ratio between two numbers that can be
negative flips sign for reasons that have nothing to do with biology.

## Summary

| Question | Function | Unit |
|---|---|---|
| Did a population change size? | `stats_group_comparison()` | sample |
| The same, after removing the compositional constraint | `clr_frequencies()` and `compositional_concordance()` | sample |
| Did a marker change level inside a population? | `stats_marker_state()` | sample |
| Is it really age or sex? | `stats_confounding()` | sample |
| Is it really the batch? | `batch_mixing_report()` | cell, descriptive |
| Is it really the gate? | `stats_threshold_drift()`, `cluster_gate_agreement()` | sample and cell |
| Which markers shift within a subcluster? | `stats_subcluster_shifts()` | cell, no p-value |
