---
title: "Statistics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(1)
```

```{r setup}
library(cyRAVEN)
```

Every test here trades assumptions against power. This article states which trade
was made at each point, and the sample size at which it should be revisited.

## 1. Aggregation

### 1.1 Pseudoreplication

The event count in an FCS file is determined by acquisition duration, not by
study design. Treating events as independent observations sets the degrees of
freedom from an operational parameter: sixteen donors contributing one million
events each yield a nominal *n* of 1.6 × 10<sup>7</sup>.

Two consequences follow. Statistical power becomes a function of instrument time,
so extending acquisition on one tube increases apparent significance. More
seriously, a single donor acquired to greater depth contributes disproportionate
weight, and inter-individual variation in that donor is indistinguishable from a
group effect.

### 1.2 Implementation

Every test aggregates to one value per sample per population before estimation,
following Weber et al. (2019, *Commun Biol* 2:183). The degrees of freedom
reflect the number of donors.

```{r ds}
mfi <- expand.grid(sample_id = sprintf("S%02d", 1:12),
                   population = "CD4 T cells", marker = "HLA-DR",
                   stringsAsFactors = FALSE)
mfi$n_cells <- 500
mfi$median_asinh <- rnorm(12, 2, 0.3) + rep(c(0, 0.8), each = 6)
mfi$pct_positive <- pmin(100, pmax(0, rnorm(12, 30, 5) + rep(c(0, 10), each = 6)))
mfi$is_control <- FALSE; mfi$qc_status <- "pass"
grp <- setNames(rep(c("HC", "Patient"), each = 6), sprintf("S%02d", 1:12))

res <- stats_marker_state(mfi, grp, reference = "HC")
res[, c("marker", "measure", "n_reference", "n_comparison", "delta_median", "p_value")]
```

`n_reference` is 6, the number of control donors, not 3,000, the number of events
they contributed.

Event-level quantities remain available through `stats_subcluster_shifts()`,
which reports effect sizes without p-values. They are hypothesis-generating.

## 2. Test selection

Group comparisons use Wilcoxon rank-sum and Kruskal-Wallis, which assume
independence and ordinal comparability.

The principal alternative, implemented in diffcyt, is limma's moderated *t*,
which stabilises variance estimates through an empirical Bayes prior shared
across markers. At mass cytometry marker counts of 30 to 40 and designed sample
sizes, this is the more powerful choice.

At a dozen markers and single-digit donors per group the prior is estimated from
too few markers to stabilise, and the normality assumption becomes load-bearing
at precisely the sample size where it cannot be assessed.

**Cost.** Reduced power under normality, and no direct mechanism for covariate
adjustment.

**Revisit at.** Approximately fifteen samples per group, where the prior becomes
informative and mixed models can accommodate repeated measures.

## 3. Compositionality

### 3.1 Constraint

Population percentages within a sample are constrained to sum to 100 and cannot
vary independently. A genuine granulocyte expansion mechanically depresses every
lymphocyte percentage. Those depressions test significant while absolute
lymphocyte counts per microlitre remain unchanged.

### 3.2 Log-ratio

`clr_frequencies()` applies the centred log-ratio, expressing each population
relative to the geometric mean of the sample composition rather than to a fixed
total.

```{r clr}
freq <- data.frame(
  sample_id = rep(sprintf("S%02d", 1:6), each = 3),
  population = rep(c("Granulocytes", "CD4 T cells", "B cells"), 6),
  pct_of_cd45_pos = as.vector(replicate(6, {v <- c(60, 25, 15) * exp(rnorm(3, 0, .1)); 100*v/sum(v)})),
  is_control = FALSE, qc_status = "pass")
head(clr_frequencies(freq)[, c("sample_id", "population", "pct_of_cd45_pos", "clr")], 3)
```

### 3.3 Concordance

Both parameterisations are tested and `compositional_concordance()` classifies
each result.

`robust_to_composition` survives both and supports interpretation as an
independent change.

`raw_only__possible_composition_artefact` is significant on percentages and not
on log-ratios, which is the configuration produced by section 3.1.

`clr_only__was_masked_by_composition` is the converse, an independent change
obscured by the constraint.

### 3.4 Limit

The log-ratio does not recover absolute abundance. Proportional expansion of
every population leaves the composition invariant and is undetectable in
frequency data by construction. Discriminating expansion from relative expansion
requires cells per microlitre, so `abundance_measure()` uses absolute counts
wherever the patient table supplies a leukocyte count.

## 4. Covariates

Age and sex are diagnosed rather than adjusted.

With eight samples per group, a model estimating age and sex effects alongside
the group term consumes the residual degrees of freedom required for the
comparison of interest. Where the covariate is strongly associated with group,
the parameters are not separable at any sample size, because the design contains
no observations at the covariate levels required to estimate the effects
independently.

`stats_confounding()` therefore reports the two conditions that jointly define a
confounder: differential distribution across groups, and association with the
outcome. `stats_rank_ancova()` fits the adjusted model on request, labels every
row `EXPLORATORY`, and records `NOT FITTED` with the reason where residual
degrees of freedom are insufficient.

## 5. Batch effects

Correction methods displace cells until batch labels mix. This is valid when
batch is independent of the study design.

Clinical cohorts rarely satisfy that condition: recruitment and acquisition are
typically sequential, so batch and group approach collinearity. Displacing cells
until batches mix then also removes the biological contrast, and the algorithm
cannot distinguish the two contributions.

`batch_mixing_report()` returns both required quantities before any correction:
iLISI against a permutation null, establishing whether batch structure exists,
and Cramér's *V*, establishing whether it is separable from group. A high *V*
means no setting makes correction safe, which is the most informative result the
diagnostic can return.

`--correct-batch` performs the correction and is refused above a configurable
threshold.

## 6. Multiplicity

Benjamini-Hochberg correction is applied within each test family, with raw and
adjusted p-values both reported.

Differential state adjusts within each measure rather than across both. Median
intensity and percent positive are two summaries of the same events and are
strongly correlated; pooling them would inflate the family with non-independent
hypotheses and over-penalise every result.

## 7. Effect sizes

Every p-value is reported with an effect size: Cliff's delta across samples, or
matched-pairs rank-biserial for paired designs.

Differences rather than ratios are reported for transformed intensities. Arcsinh
and logicle scales admit negative values, on which a ratio is undefined in
interpretation and changes sign without a corresponding change in biology.

## 8. Measurement resolution

An effect size states how large a difference is relative to the spread between
donors. It does not state whether the instrument and the gating strategy could
resolve a difference of that size at all.

`difference_over_gate_u` supplies the second quantity: the observed difference in
medians divided by the typical within-sample uncertainty on the frequency, itself
propagated from where the thresholds were placed. Below one, the groups differ by
less than the distance the cut travels under resampling of the cells that
determined it.

This is a screen, not a test. The uncertainty is partly shared across the run,
since every sample passes through one panel, one transform and one placement
rule, so it cancels in part when a difference is taken and the ratio is
conservative. A ratio below one is a reason to open `threshold_uncertainty.csv`
before interpreting the comparison, not grounds to discard it.

The order to read remains: adjusted p, then effect size, then this. A result that
survives all three is one where the groups differ, the difference is large
relative to donor variation, and it is larger than the gate can move on its own.

## 9. Summary

| Question | Function | Unit |
|---|---|---|
| Did population abundance change? | `stats_group_comparison()` | sample |
| Does it survive the compositional constraint? | `clr_frequencies()`, `compositional_concordance()` | sample |
| Did marker expression change within a population? | `stats_marker_state()` | sample |
| Is the contrast confounded by age or sex? | `stats_confounding()` | sample |
| Is the contrast confounded by batch? | `batch_mixing_report()` | cell, descriptive |
| Is the contrast an artefact of the gate? | `stats_threshold_drift()`, `cluster_gate_agreement()` | sample and cell |
| Is the difference bigger than the gate's own uncertainty? | `run_gate_uncertainty()`, `annotate_gate_uncertainty()` | sample |
| Which markers shift within a subcluster? | `stats_subcluster_shifts()` | cell, no p-value |
